---
title: "Introduction to scKnockoff - Easy"
author: "Hyunjae Lee"
output: github_document
vignette: >
  %\VignetteIndexEntry{Introduction to scKnockoff - Easy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# 1. Load scKnockoff

```{r}
# if (!requireNamespace("scKnockoff", quietly = TRUE)) {
#   devtools::install_github("HJLee196/scKnockoff")
# }

library(scKnockoff)
```

# 2. Create dataset

For convenience, the package provides `make_toy_seurat_ad()`, which generates a toy Seurat object under a Gaussian latent-factor model. The synthetic single-cell RNA-seq dataset is generated from

\[
G = X B_0 + A B_1 + E,
\]

where \(G\) represents the latent Gaussian log-expression matrix, \(X\) contains observed covariates such as disease status, batch, age, and cell type, \(A\) contains Gaussian latent factors, and \(E\) is Gaussian noise. Disease status is assigned at the donor level, and nonzero disease effects are assigned to a subset of genes in \(B_0\), defining the ground-truth signals. To construct a Seurat object, the latent log-mean expression matrix \(G\) is converted to a positive mean expression matrix by setting \(\mu_{ij} = \exp(G_{ij})\), and observed UMI-like counts are generated from a Poisson model.

```{r}

Seurat_toy = make_toy_seurat_ad(n_genes = 100, 
                                donors_per_group = 6,
                                n_signals = 50,
                                # fold-change on relative abundance scale
                                signal_strength = 2,
                                # latent dimension
                                r = 5,
                                seed = 1)

```

# 3. Add the cellular detection rate (CDR) as a latent variable in the toy dataset.

In addition to `batch`, `cell_type`, and `age`, we compute and include the cellular detection rate (CDR) as a latent variable. Here, `CDR` summarizes the overall detection level of each cell and is commonly used to adjust for cell-level technical variation in single-cell RNA-seq data.

```{r}
feature.names <- rownames(Seurat_toy)

np_data.count <- Seurat::GetAssayData(object = Seurat_toy, layer = "count")

np_data.matrix.exp <- Matrix::t(np_data.count > 0) # indicator matrix for expressed genes,
np_data.matrix.exp.count <- Matrix::colSums(np_data.matrix.exp)

# Calculate cellular detection rate (CDR)
np_data.matrix <- np_data.count # do not transpose here. Keep the genes as rows.

CDR <- Matrix::rowMeans(np_data.matrix.exp) # expressed genes in each CELL
Seurat_toy$CDR <- CDR
```

# 4. Select the significant genes using `full_process`

We then apply `full_process()` to identify genes associated with disease status. In this example, we compare cells from the AD group against cells from the Control group by setting `ident.1 = "AD"` and `ident.2 = "Control"`. Setting `PC = NULL` allows the number of principal components to be estimated automatically using the bulk eigenvalue matching analysis (BEMA) method [@ke2023]. The covariates `batch`, `cell_type`, `age`, and `CDR` are included in both `latent.vars_imp` and `latent.vars_comp` to adjust for potential confounding effects during the imputation and comparison steps. 

We use `test.use = "LCD"` to construct feature-importance statistics based on the Lasso coefficient difference statistic, and `m = 1` specifies that a single knockoff copy is generated for each gene.

```{r}
result_LCD = 
  full_process(Seurat_data = Seurat_toy,
               PC = NULL,
               ident.1 = "AD",
               ident.2 = "Control",
               latent.vars_imp = c("batch", "cell_type", "age", "CDR"),
               latent.vars_comp = c("batch", "cell_type", "age", "CDR"),
               test.use = "LCD",
               m = 1)
```

We then evaluate the false discovery proportion (FDP) and power of the selected genes.

```{r}
true_signal <- Seurat_toy@misc$true_signal

print("FDP and Power")
print((length(result_LCD$selected) - sum(result_LCD$selected %in% true_signal))/length(result_LCD$selected)) # FDR
print(sum(result_LCD$selected %in% true_signal)/length(true_signal)) # Power

```

The feature-importance statistics can also be constructed using other testing methods supported by `Seurat::FindMarkers()`. In addition, setting `m > 1` enables the use of multiple knockoffs. In the following example, we use `test.use = "MAST"` and generate five knockoff copies for each gene by setting `m = 5`. Here, we set `PC = 5`, corresponding to the true number of latent factors used to generate the toy dataset.

```{r}
result_MAST = 
  full_process(Seurat_data = Seurat_toy,
               PC = 5,
               ident.1 = "AD",
               ident.2 = "Control",
               latent.vars_imp = c("batch", "cell_type", "age", "CDR"),
               latent.vars_comp = c("batch", "cell_type", "age", "CDR"),
               test.use = "MAST",
               m = 5)
```

We then evaluate the false discovery proportion (FDP) and power of the selected genes.

```{r}
print("FDP and Power")
print((length(result_MAST$selected) - sum(result_MAST$selected %in% true_signal))/length(result_MAST$selected)) # FDR
print(sum(result_MAST$selected %in% true_signal)/length(true_signal)) # Power

```

The ground-truth signal genes and the genes selected by LCD and MAST are shown below:

```{r}
cat("Ground-truth signal genes:\n")
cat(Seurat_toy@misc$true_signal_genes, sep = ", ")
cat("\n\n")

cat("Genes selected by LCD (m=1):\n")
cat(result_LCD$variables_name, sep = ", ")
cat("\n\n")

cat("Genes selected by MAST (m=5):\n")
cat(result_MAST$variables_name, sep = ", ")
cat("\n")

```

