---
title: "Introduction to scKnockoff"
author: "Hyunjae Lee"
output: github_document
vignette: >
  %\VignetteIndexEntry{Introduction to scKnockoff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# 1. Load scKnockoff

```{r}
# if (!requireNamespace("scKnockoff", quietly = TRUE)) {
#   devtools::install_github("HJLee196/scKnockoff")
# }

library(scKnockoff)
```

# 2. Create dataset

We generate a synthetic single-cell RNA-seq dataset from the latent factor model \(G = X B_0 + A B_1 + E\), where \(X\) contains observed covariates such as disease status, batch, age, and cell type. Disease status is assigned at the donor level, and a subset of genes is assigned nonzero disease effects in \(B_0\), serving as the ground-truth signals. The latent expression matrix \(G\) is transformed into cell-specific relative expression probabilities, from which observed UMI counts are sampled using a multinomial model.

```{r}
make_toy_seurat_ad <- function(
    n_genes = 1000,
    donors_per_group = 6,
    cell_types = c("Astro", "Micro", "Neuron"),
    cells_per_donor_per_type = 50,
    n_batch = 2,
    # disease-associated genes
    n_signals = 50,
    # fold-change on relative abundance scale
    signal_strength = 2,
    # latent dimension
    r = 5,
    # noise level in Gaussian latent layer
    sigma = 0.5,
    # library size distribution
    mean_umi = 5000,
    lib_size_sd = 0.4,
    # effect sizes for observed covariates
    age_effect_sd = 0.05,
    batch_effect_sd = 0.15,
    cell_type_effect_sd = 0.25,
    # loading sparsity
    loading_prop = 0.2,
    loading_sd = 0.7,
    normalize = TRUE,
    scale_data = TRUE,
    seed = 1
) {
  if (!requireNamespace("Seurat", quietly = TRUE)) {
    stop("Seurat is required to build the toy Seurat object. Install Seurat or skip this example.")
  }

  if (!requireNamespace("Matrix", quietly = TRUE)) {
    stop("Matrix is required to create a sparse count matrix.")
  }

  set.seed(seed)

  # 1. Metadata

  n_types <- length(cell_types)
  n_donors <- 2 * donors_per_group
  donors <- paste0("D", seq_len(n_donors))

  donor_disease <- rep(c("Control", "AD"), each = donors_per_group)
  names(donor_disease) <- donors

  batch_levels <- paste0("B", seq_len(n_batch))
  batch_by_donor <- sample(batch_levels, size = n_donors, replace = TRUE)
  names(batch_by_donor) <- donors

  donor_age <- round(runif(n_donors, 65, 90))
  names(donor_age) <- donors

  n_cells <- n_donors * n_types * cells_per_donor_per_type

  meta <- data.frame(
    cell_id = paste0("cell", seq_len(n_cells)),
    donor = rep(donors, each = n_types * cells_per_donor_per_type),
    cell_type = rep(rep(cell_types, each = cells_per_donor_per_type), times = n_donors),
    stringsAsFactors = FALSE
  )

  meta$disease <- unname(donor_disease[meta$donor])
  meta$batch <- unname(batch_by_donor[meta$donor])
  meta$age <- unname(donor_age[meta$donor])

  meta$disease <- factor(meta$disease, levels = c("Control", "AD"))
  meta$batch <- factor(meta$batch, levels = batch_levels)
  meta$cell_type <- factor(meta$cell_type, levels = cell_types)

  genes <- paste0("gene", seq_len(n_genes))

  # 2. Observed covariates X

  # This is the X in G = X B0 + A B1 + E.
  # diseaseAD coefficient defines true disease-associated genes.

  X <- stats::model.matrix(
    ~ disease + batch + age + cell_type,
    data = meta
  )

  # Standardize age column only, if present.
  if ("age" %in% colnames(X)) {
    X[, "age"] <- as.numeric(scale(X[, "age"]))
  }

  q <- ncol(X)

  # 3. Construct B0

  B0 <- matrix(0, nrow = q, ncol = n_genes)
  rownames(B0) <- colnames(X)
  colnames(B0) <- genes

  # Baseline gene abundance through intercept.
  # This makes genes have heterogeneous baseline expression.
  baseline_log_abundance <- rnorm(n_genes, mean = 0, sd = 1)
  B0["(Intercept)", ] <- baseline_log_abundance

  # Small age effects.
  if ("age" %in% rownames(B0)) {
    B0["age", ] <- rnorm(n_genes, mean = 0, sd = age_effect_sd)
  }

  # Batch effects.
  batch_rows <- grep("^batch", rownames(B0), value = TRUE)
  if (length(batch_rows) > 0L) {
    for (rr in batch_rows) {
      B0[rr, ] <- rnorm(n_genes, mean = 0, sd = batch_effect_sd)
    }
  }

  # Cell-type effects.
  cell_type_rows <- grep("^cell_type", rownames(B0), value = TRUE)
  if (length(cell_type_rows) > 0L) {
    for (rr in cell_type_rows) {
      B0[rr, ] <- rnorm(n_genes, mean = 0, sd = cell_type_effect_sd)
    }
  }

  # Disease signals.
  signal_genes <- integer(0)
  signal_gene_names <- character(0)

  if (n_signals > 0L) {
    signal_genes <- sort(sample(seq_len(n_genes), n_signals))
    signal_gene_names <- genes[signal_genes]

    disease_row <- "diseaseAD"

    if (!disease_row %in% rownames(B0)) {
      stop("Could not find `diseaseAD` in the design matrix.", call. = FALSE)
    }

    n_up <- ceiling(n_signals / 2)
    n_down <- n_signals - n_up

    disease_effect <- c(
      rep(log(signal_strength), n_up),
      rep(-log(signal_strength), n_down)
    )

    disease_effect <- sample(disease_effect)

    B0[disease_row, signal_genes] <- disease_effect
  }

  # 4. Latent factors A and loadings B1

  A <- matrix(
    stats::rnorm(n_cells * r),
    nrow = n_cells,
    ncol = r
  )
  rownames(A) <- meta$cell_id
  colnames(A) <- paste0("LF", seq_len(r))

  B1 <- matrix(0, nrow = r, ncol = n_genes)
  rownames(B1) <- colnames(A)
  colnames(B1) <- genes

  n_loading_genes <- max(1L, floor(loading_prop * n_genes))

  for (k in seq_len(r)) {
    idx <- sample(seq_len(n_genes), n_loading_genes)
    B1[k, idx] <- stats::rnorm(length(idx), mean = 0, sd = loading_sd)
  }

  # 5. Gaussian latent expression layer

  # G = X B0 + A B1 + E

  E <- matrix(
    stats::rnorm(n_cells * n_genes, mean = 0, sd = sigma),
    nrow = n_cells,
    ncol = n_genes
  )

  G <- X %*% B0 + A %*% B1 + E
  rownames(G) <- meta$cell_id
  colnames(G) <- genes

  # 6. Library sizes and multinomial counts

  library_size <- round(
    exp(stats::rnorm(n_cells, mean = log(mean_umi), sd = lib_size_sd))
  )

  library_size[library_size < 1L] <- 1L

  counts <- matrix(
    0L,
    nrow = n_genes,
    ncol = n_cells,
    dimnames = list(genes, meta$cell_id)
  )

  for (i in seq_len(n_cells)) {
    # Stable softmax.
    eta <- G[i, ]
    eta <- eta - max(eta)

    prob <- exp(eta)
    prob <- prob / sum(prob)

    counts[, i] <- as.integer(
      stats::rmultinom(
        n = 1,
        size = library_size[i],
        prob = prob
      )
    )
  }

  counts <- Matrix::Matrix(counts, sparse = TRUE)

  # 7. Build Seurat object

  seu <- Seurat::CreateSeuratObject(
    counts = counts,
    project = "toyAD"
  )

  seu$donor <- meta$donor
  seu$disease <- meta$disease
  seu$cell_type <- meta$cell_type
  seu$batch <- meta$batch
  seu$age <- meta$age
  seu$library_size_true <- library_size

  Seurat::Idents(seu) <- "disease"

  if (isTRUE(normalize)) {
    seu <- Seurat::NormalizeData(seu, verbose = FALSE)
  }

  if (isTRUE(scale_data)) {
    seu <- Seurat::ScaleData(seu, features = rownames(seu), verbose = FALSE)
  }

  # 8. Store ground truth

  beta_disease <- rep(0, n_genes)
  names(beta_disease) <- genes

  if (n_signals > 0L) {
    beta_disease[signal_genes] <- B0["diseaseAD", signal_genes]
  }

  seu@misc$true_signal <- signal_genes
  seu@misc$true_signal_genes <- signal_gene_names
  seu@misc$beta_disease <- beta_disease
  seu@misc$B0 <- B0
  seu@misc$B1 <- B1
  seu@misc$A <- A
  seu@misc$X <- X
  seu@misc$G <- G
  seu@misc$library_size_true <- library_size
  seu@misc$simulation_model <- "G = X B0 + A B1 + E; counts_i ~ Multinomial(N_i, softmax(G_i))"

  seu
}

Seurat_toy = make_toy_seurat_ad(n_genes = 100, 
                                donors_per_group = 6,
                                n_signals = 50,
                                # fold-change on relative abundance scale
                                signal_strength = 2,
                                # latent dimension
                                r = 5,
                                seed = 1)

```


# 3. Imputation using sc-softImpute

In addition to `batch`, `cell_type`, and `age`, we compute and include the cellular detection rate (CDR) as a latent variable.

```{r}
feature.names <- rownames(Seurat_toy)

np_data.count <- Seurat::GetAssayData(object = Seurat_toy, layer = "count")

np_data.matrix.exp <- Matrix::t(np_data.count > 0) # indicator matrix for expressed genes,
np_data.matrix.exp.count <- Matrix::colSums(np_data.matrix.exp)

# Calculate cellular detection rate (CDR)
np_data.matrix <- np_data.count # do not transpose here. Keep the genes as rows.

CDR <- Matrix::rowMeans(np_data.matrix.exp) # expressed genes in each CELL
Seurat_toy$CDR <- CDR
```

Before applying `sc_softImpute`, we first need to create a matrix of the latent variables.

```{r}
latent.vars_imp = c("batch", "cell_type", "age", "CDR")
# generate X
X <- NULL
for (imp_vars_temp in latent.vars_imp) {

  # meta.data column as vector
  v <- Seurat_toy[[imp_vars_temp]][, 1, drop = TRUE]

  # if non-numeric: one-hot encode
  if (!is.numeric(v)) {
    mm <- stats::model.matrix(~ x, data = data.frame(x = v))  # includes intercept
    mm <- mm[, -1, drop = FALSE]  # drop intercept -> k-1 dummies
    # optional: prefix colnames to avoid collisions
    colnames(mm) <- paste0(imp_vars_temp, ":", colnames(mm))
    vmat <- mm
  } else {
    vmat <- matrix(v, ncol = 1)
    colnames(vmat) <- imp_vars_temp
  }

  X <- if (is.null(X)) vmat else cbind(X, vmat)
}
```

We also filter out genes that are expressed in only a few cells to satisfy the model assumptions. In this example, we set `PC = 5`.

```{r}
PC = 5

# Exclude some genes
if (is.null(X)){
  X_p <- 0
} else {
  X_p <- ncol(X)
}
feature.exp.ind <- which(np_data.matrix.exp.count > (PC + X_p + 1)) 

feature.names.new <- feature.names[feature.exp.ind]  

np_data.matrix.exp <- np_data.matrix.exp[,feature.exp.ind]
np_data.matrix.exp.count <- np_data.matrix.exp.count[feature.exp.ind]

# Refresh Seurat_toy
Seurat_toy <- subset(x = Seurat_toy, features = feature.names.new)
```

To apply `sc_softImpute`, we center the expressed part of the data matrix.

```{r}
np_data.matrix <- Seurat::GetAssayData(object = Seurat_toy, layer = "data") # Normalized data matrix
np_data.matrix <- Matrix::t(np_data.matrix)

# Center target matrix, but only center the expressed parts.
np_data.matrix.fix <- (Matrix::colSums(np_data.matrix*np_data.matrix.exp))/np_data.matrix.exp.count # na.rm = T is removed
np_data.matrix.center <- methods::as(sweep(np_data.matrix,2,np_data.matrix.fix,FUN = "-")*(np_data.matrix.exp), Class = "dgCMatrix")
```

We now apply `sc_softImpute`. Note that `np_data.matrix.exp` should be a numeric matrix with the same dimensions as `np_data.matrix`, where entries are 1 for observed values and 0 for missing values.

```{r}
#### Imputation ####
np_data.matrix.exp <- np_data.matrix.exp*1
np_data.matrix.imp.list <- sc_softImpute(np_data.matrix = np_data.matrix.center,
                                         np_data.matrix.exp = np_data.matrix.exp,
                                         X = X,
                                         PC = PC,
                                         max_it = 100)
```

# 4. Generating Knockoffs for each method

In scKnockoff, we provide four knockoff generation methods:
- LR knockoff
- Multi-LR knockoff
- Decomp knockoff
- Multi-decomp knockoff

```{r}
lr_knock <- create_lr_knock(np_data.matrix.imp = np_data.matrix.imp.list$X_imp,
                            Xl = np_data.matrix.imp.list$Xl,
                            Bl = np_data.matrix.imp.list$Bl,
                            err = np_data.matrix.imp.list$err)

decomp_knock <- create_decomp_knock(np_data.matrix.imp = np_data.matrix.imp.list$X_imp,
                                    Xl = np_data.matrix.imp.list$Xl,
                                    Bl = np_data.matrix.imp.list$Bl,
                                    err = np_data.matrix.imp.list$err,
                                    PC = np_data.matrix.imp.list$PC, 
                                    decomp = TRUE)

m <- 5 # Number of knockoff copies
multi_lr_knock <- create_multi_lr_knock(np_data.matrix.imp = np_data.matrix.imp.list$X_imp,
                                        Xl = np_data.matrix.imp.list$Xl,
                                        Bl = np_data.matrix.imp.list$Bl,
                                        err = np_data.matrix.imp.list$err,
                                        m = m)

multi_decomp_knock <- create_multi_decomp_knock(np_data.matrix.imp = np_data.matrix.imp.list$X_imp,
                                                Xl = np_data.matrix.imp.list$Xl,
                                                Bl = np_data.matrix.imp.list$Bl,
                                                err = np_data.matrix.imp.list$err,
                                                PC = np_data.matrix.imp.list$PC, 
                                                m = m)
```

After generating the knockoffs, we rescale them.

```{r}
lr_knock <- rescale_knockoff(X_knk = lr_knock, 
                             np_data.matrix.exp = np_data.matrix.exp, 
                             center = np_data.matrix.fix)

decomp_knock <- rescale_knockoff(X_knk = decomp_knock, 
                                 np_data.matrix.exp = np_data.matrix.exp, 
                                 center = np_data.matrix.fix)

multi_lr_knock <- rescale_knockoff(X_knk = multi_lr_knock, 
                                   np_data.matrix.exp = np_data.matrix.exp, 
                                   center = np_data.matrix.fix)

multi_decomp_knock <- rescale_knockoff(X_knk = multi_decomp_knock, 
                                       np_data.matrix.exp = np_data.matrix.exp, 
                                       center = np_data.matrix.fix)
```

# 5. Compute the feature importance for each knockoff

We compute feature importance using testing procedures provided by `Seurat::FindMarkers`, including the lasso coefficient difference (LCD) statistic, which can be specified by setting `test.use = "LCD"` in `feature_importance`. In our implementation, we use `test.use = "LCD"` for the LR and decomp knockoffs, and `test.use = "LR"` for the multi-LR and multi-decomp knockoffs.

```{r}
latent.vars_comp = latent.vars_imp

W_imp_lr <- feature_importance(np_data.sub = Seurat_toy,
                               np_data.knockoff = lr_knock,
                               bonf = TRUE,
                               ident.1 = "Control", ident.2 = "AD", slot_name = "data",
                               test.use = "LCD", latent.vars = latent.vars_comp)

W_imp_decomp <- feature_importance(np_data.sub = Seurat_toy,
                                   np_data.knockoff = decomp_knock,
                                   bonf = TRUE,
                                   ident.1 = "Control", ident.2 = "AD", slot_name = "data",
                                   test.use = "LCD", latent.vars = latent.vars_comp)


W_imp_multi_lr <- feature_importance(np_data.sub = Seurat_toy,
                                     np_data.knockoff = multi_lr_knock,
                                     bonf = TRUE,
                                     ident.1 = "Control", ident.2 = "AD", slot_name = "data",
                                     test.use = "LR", latent.vars = latent.vars_comp)

W_imp_multi_decomp <- feature_importance(np_data.sub = Seurat_toy,
                                         np_data.knockoff = multi_decomp_knock,
                                         bonf = TRUE,
                                         ident.1 = "Control", ident.2 = "AD", slot_name = "data",
                                         test.use = "LR", latent.vars = latent.vars_comp)

```

Using `knockfilter_select`, we select potentially significant genes at a target false discovery rate of `q = 0.1`.

```{r}
selected_lr <- knockfilter_select(W_imp = W_imp_lr, q = 0.1)
selected_decomp <- knockfilter_select(W_imp = W_imp_decomp, q = 0.1)
selected_multi_lr <- knockfilter_select(W_imp = W_imp_multi_lr, q = 0.1)
selected_multi_decomp <- knockfilter_select(W_imp = W_imp_multi_decomp, q = 0.1)
```

The eBH procedure can also be applied using `ebh_select` to select significant genes.

```{r}
selected_e_lr <- ebh_select(W_imp = W_imp_multi_lr, q = 0.1)
selected_e_decomp <- ebh_select(W_imp = W_imp_multi_decomp, q = 0.1)
```

We then evaluate the false discovery proportions (FDPs) and powers of the selected genes.

```{r}
true_signal <- Seurat_toy@misc$true_signal

print("LR knockoff - LCD")
print((length(selected_lr) - sum(selected_lr %in% true_signal))/length(selected_lr)) # FDP
print(sum(selected_lr %in% true_signal)/length(true_signal)) # Power

print("Decomp knockoff - LCD")
print((length(selected_decomp) - sum(selected_decomp %in% true_signal))/length(selected_decomp)) # FDP
print(sum(selected_decomp %in% true_signal)/length(true_signal)) # Power

print("Multi-LR knockoff - LR")
print((length(selected_multi_lr) - sum(selected_multi_lr %in% true_signal))/length(selected_multi_lr)) # FDP
print(sum(selected_multi_lr %in% true_signal)/length(true_signal)) # Power

print("Multi-decomp knockoff - LR")
print((length(selected_multi_decomp) - sum(selected_multi_decomp %in% true_signal))/length(selected_multi_decomp)) # FDP
print(sum(selected_multi_decomp %in% true_signal)/length(true_signal)) # Power

print("e-LR knockoff - LR")
print((length(selected_e_lr) - sum(selected_e_lr %in% true_signal))/length(selected_e_lr)) # FDP
print(sum(selected_e_lr %in% true_signal)/length(true_signal)) # Power

print("e-decomp knockoff - LR")
print((length(selected_e_decomp) - sum(selected_e_decomp %in% true_signal))/length(selected_e_decomp)) # FDP
print(sum(selected_e_decomp %in% true_signal)/length(true_signal)) # Power
```




