---
title: "Introduction to scKnockoff: Step-by-Step Customization"
output: github_document
vignette: >
  %\VignetteIndexEntry{Introduction to scKnockoff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates a step-by-step workflow for applying `scKnockoff` to a toy single-cell RNA-seq dataset. We illustrate data generation, imputation using `sc_softImpute`, knockoff construction, feature-importance calculation, and gene selection.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# 1. Load scKnockoff

```{r}
# if (!requireNamespace("scKnockoff", quietly = TRUE)) {
#   devtools::install_github("HJLee196/scKnockoff")
# }

library(scKnockoff)
```

# 2. Create Dataset

For convenience, the package provides `make_toy_seurat_ad()`, which generates a toy Seurat object under a Gaussian latent-factor model. The synthetic single-cell RNA-seq dataset is generated from

\[
G = X B_0 + A B_1 + E,
\]

where \(G\) represents the latent Gaussian log-expression matrix, \(X\) contains observed covariates such as disease status, batch, age, and cell type, \(A\) contains Gaussian latent factors, and \(E\) is Gaussian noise. Disease status is assigned at the donor level, and nonzero disease effects are assigned to a subset of genes in \(B_0\), defining the ground-truth signals. To construct a Seurat object, the latent log-mean expression matrix \(G\) is converted to a positive mean expression matrix by setting \(\mu_{ij} = \exp(G_{ij})\), and observed UMI-like counts are generated from a Poisson model.

```{r}
Seurat_toy = make_toy_seurat_ad(n_genes = 200, 
                                donors_per_group = 6,
                                n_signals = 50,
                                # fold-change on relative abundance scale
                                signal_strength = 2,
                                # latent dimension
                                r = 5,
                                # standard deviation of noise
                                sigma = 2,
                                seed = 2)

```


# 3. Imputation Using sc-softImpute

In addition to `batch`, `cell_type`, and `age`, we compute and include the cellular detection rate (CDR) as an observed covariate.

```{r}
feature.names <- rownames(Seurat_toy)

np_data.count <- Seurat::GetAssayData(object = Seurat_toy, layer = "count")

# Cell-by-gene indicator matrix: 1 if a gene is expressed in a cell.
np_data.matrix.exp <- Matrix::t(np_data.count > 0)
# Number of cells in which each gene is expressed.
np_data.matrix.exp.count <- Matrix::colSums(np_data.matrix.exp)

# Calculate cellular detection rate (CDR)
# Fraction of expressed genes in each cell.
CDR <- Matrix::rowMeans(np_data.matrix.exp)

Seurat_toy$CDR <- CDR
```

Before applying `sc_softImpute`, we first construct a covariate matrix from the variables used for imputation.

```{r}
latent.vars_imp = c("batch", "cell_type", "age", "CDR")
# generate X
X <- NULL
for (imp_vars_temp in latent.vars_imp) {

  # meta.data column as vector
  v <- Seurat_toy[[imp_vars_temp]][, 1, drop = TRUE]

  # if non-numeric: one-hot encode
  if (!is.numeric(v)) {
    mm <- stats::model.matrix(~ x, data = data.frame(x = v))  # includes intercept
    mm <- mm[, -1, drop = FALSE]  # drop intercept -> k-1 dummies
    # optional: prefix colnames to avoid collisions
    colnames(mm) <- paste0(imp_vars_temp, ":", colnames(mm))
    vmat <- mm
  } else {
    vmat <- matrix(v, ncol = 1)
    colnames(vmat) <- imp_vars_temp
  }

  X <- if (is.null(X)) vmat else cbind(X, vmat)
}
```

We also filter out genes that are expressed in too few cells. This ensures that each retained gene has enough observed entries relative to the number of covariates and the chosen latent rank. In this example, we set `PC = 5`.

```{r}
PC = 5

# Exclude some genes
if (is.null(X)){
  X_p <- 0
} else {
  X_p <- ncol(X)
}
feature.exp.ind <- which(np_data.matrix.exp.count > (PC + X_p + 1)) 

feature.names.new <- feature.names[feature.exp.ind]  

np_data.matrix.exp <- np_data.matrix.exp[,feature.exp.ind]
np_data.matrix.exp.count <- np_data.matrix.exp.count[feature.exp.ind]

# Refresh Seurat_toy
Seurat_toy <- subset(x = Seurat_toy, features = feature.names.new)
```

To apply `sc_softImpute`, we center the expressed part of the data matrix.

```{r}
np_data.matrix <- Seurat::GetAssayData(object = Seurat_toy, layer = "data") # Normalized data matrix
np_data.matrix <- Matrix::t(np_data.matrix)

# Center target matrix, but only center the expressed parts.
np_data.matrix.fix <- (Matrix::colSums(np_data.matrix*np_data.matrix.exp))/np_data.matrix.exp.count # na.rm = T is removed
np_data.matrix.center <- methods::as(sweep(np_data.matrix,2,np_data.matrix.fix,FUN = "-")*(np_data.matrix.exp), Class = "dgCMatrix")
```

We now apply `sc_softImpute`. Note that `np_data.matrix.exp` should be a numeric matrix with the same dimensions as `np_data.matrix`, where entries are 1 for observed values and 0 for missing values.

```{r}
#### Imputation ####
np_data.matrix.exp <- np_data.matrix.exp*1
np_data.matrix.imp.list <- sc_softImpute(np_data.matrix = np_data.matrix.center,
                                         np_data.matrix.exp = np_data.matrix.exp,
                                         X = X,
                                         PC = PC,
                                         max_it = 100)
```

# 4. Knockoff Generation Methods

In scKnockoff, we provide four knockoff generation methods:

- LR knockoff
- Multi-LR knockoff
- Decomp knockoff
- Multi-decomp knockoff

```{r}
lr_knock <- create_lr_knock(np_data.matrix.imp = np_data.matrix.imp.list$X_imp,
                            Xl = np_data.matrix.imp.list$Xl,
                            Bl = np_data.matrix.imp.list$Bl,
                            err = np_data.matrix.imp.list$err)

decomp_knock <- create_decomp_knock(np_data.matrix.imp = np_data.matrix.imp.list$X_imp,
                                    Xl = np_data.matrix.imp.list$Xl,
                                    Bl = np_data.matrix.imp.list$Bl,
                                    err = np_data.matrix.imp.list$err,
                                    PC = np_data.matrix.imp.list$PC, 
                                    decomp = TRUE)

m <- 5 # Number of knockoff copies
multi_lr_knock <- create_multi_lr_knock(np_data.matrix.imp = np_data.matrix.imp.list$X_imp,
                                        Xl = np_data.matrix.imp.list$Xl,
                                        Bl = np_data.matrix.imp.list$Bl,
                                        err = np_data.matrix.imp.list$err,
                                        m = m)

multi_decomp_knock <- create_multi_decomp_knock(np_data.matrix.imp = np_data.matrix.imp.list$X_imp,
                                                Xl = np_data.matrix.imp.list$Xl,
                                                Bl = np_data.matrix.imp.list$Bl,
                                                err = np_data.matrix.imp.list$err,
                                                PC = np_data.matrix.imp.list$PC, 
                                                m = m)
```

After generating the knockoffs, we rescale them.

```{r}
lr_knock <- rescale_knockoff(X_knk = lr_knock, 
                             np_data.matrix.exp = np_data.matrix.exp, 
                             center = np_data.matrix.fix)

decomp_knock <- rescale_knockoff(X_knk = decomp_knock, 
                                 np_data.matrix.exp = np_data.matrix.exp, 
                                 center = np_data.matrix.fix)

multi_lr_knock <- rescale_knockoff(X_knk = multi_lr_knock, 
                                   np_data.matrix.exp = np_data.matrix.exp, 
                                   center = np_data.matrix.fix)

multi_decomp_knock <- rescale_knockoff(X_knk = multi_decomp_knock, 
                                       np_data.matrix.exp = np_data.matrix.exp, 
                                       center = np_data.matrix.fix)
```

# 5. Compute the Feature Importance for Each Knockoff

We compute feature-importance statistics using `feature_importance()`. The function uses `Seurat::FindMarkers` for supported testing procedures and additionally implements the lasso coefficient difference (LCD) statistic, which is not provided by `Seurat::FindMarkers`. The LCD statistic can be selected by setting `test.use = "LCD"`. Here, we set `latent.vars_comp = latent.vars_imp` so that the same covariates used during imputation are also adjusted for when computing the feature-importance statistics.

In our implementation, we use `test.use = "LCD"` for the LR and decomp knockoffs, and `test.use = "LR"` for the multi-LR and multi-decomp knockoffs. When `bonf = TRUE`, the feature-importance statistics are computed using the Bonferroni-adjusted p-values returned by `Seurat::FindMarkers`, rather than the unadjusted p-values.

```{r}
latent.vars_comp = latent.vars_imp

W_imp_lr <- feature_importance(np_data.sub = Seurat_toy,
                               np_data.knockoff = lr_knock,
                               bonf = TRUE,
                               ident.1 = "Control", ident.2 = "AD", slot_name = "data",
                               test.use = "LCD", latent.vars = latent.vars_comp)

W_imp_decomp <- feature_importance(np_data.sub = Seurat_toy,
                                   np_data.knockoff = decomp_knock,
                                   bonf = TRUE,
                                   ident.1 = "Control", ident.2 = "AD", slot_name = "data",
                                   test.use = "LCD", latent.vars = latent.vars_comp)


W_imp_multi_lr <- feature_importance(np_data.sub = Seurat_toy,
                                     np_data.knockoff = multi_lr_knock,
                                     bonf = TRUE,
                                     ident.1 = "Control", ident.2 = "AD", slot_name = "data",
                                     test.use = "LR", latent.vars = latent.vars_comp)

W_imp_multi_decomp <- feature_importance(np_data.sub = Seurat_toy,
                                         np_data.knockoff = multi_decomp_knock,
                                         bonf = TRUE,
                                         ident.1 = "Control", ident.2 = "AD", slot_name = "data",
                                         test.use = "LR", latent.vars = latent.vars_comp)

```

Using `knockfilter_select`, we select potentially significant genes at a target false discovery rate of `q = 0.1`.

```{r}
selected_lr <- knockfilter_select(W_imp = W_imp_lr, q = 0.1)
selected_decomp <- knockfilter_select(W_imp = W_imp_decomp, q = 0.1)
selected_multi_lr <- knockfilter_select(W_imp = W_imp_multi_lr, q = 0.1)
selected_multi_decomp <- knockfilter_select(W_imp = W_imp_multi_decomp, q = 0.1)
```

The eBH procedure can also be applied using `ebh_select` to select significant genes.

```{r}
selected_e_lr <- ebh_select(W_imp = W_imp_multi_lr, q = 0.1)
selected_e_decomp <- ebh_select(W_imp = W_imp_multi_decomp, q = 0.1)
```

We then evaluate the false discovery proportions (FDPs) and powers of the selected genes.

```{r}
true_signal <- Seurat_toy@misc$true_signal

compute_fdp <- function(selected, true_signal) {
  if (length(selected) == 0) return(0)
  (length(selected) - sum(selected %in% true_signal)) / length(selected)
}

compute_power <- function(selected, true_signal) {
  sum(selected %in% true_signal) / length(true_signal)
}

results <- data.frame(
  Method = c("LR knockoff - LCD",
             "Decomp knockoff - LCD",
             "Multi-LR knockoff - LR",
             "Multi-decomp knockoff - LR",
             "e-LR knockoff - LR",
             "e-decomp knockoff - LR"),
  FDP = c(compute_fdp(selected_lr, true_signal),
          compute_fdp(selected_decomp, true_signal),
          compute_fdp(selected_multi_lr, true_signal),
          compute_fdp(selected_multi_decomp, true_signal),
          compute_fdp(selected_e_lr, true_signal),
          compute_fdp(selected_e_decomp, true_signal)),
  Power = c(compute_power(selected_lr, true_signal),
            compute_power(selected_decomp, true_signal),
            compute_power(selected_multi_lr, true_signal),
            compute_power(selected_multi_decomp, true_signal),
            compute_power(selected_e_lr, true_signal),
            compute_power(selected_e_decomp, true_signal))
)

results

```


